---
title: "Total_parasitism_analysis"
output: html_document
---

``` {r required packages, include = FALSE}
library(arm, quietly = TRUE) # display(), coefplot() etc.
library(ggeffects, quietly = TRUE) # Estimated Marginal Means
library(ggfortify, quietly = TRUE) # autoplot()
library(marginaleffects, quietly = TRUE) # Estimated Marginal Means
library(SMPracticals, quietly = TRUE) # darwin dataframe
library("tidyverse")
library("lubridate")
library("viridis")
library(dplyr)
library(emmeans)
```

## Total parasitism analysis, MT data
Reading in the data:

``` {r reading data}
tot_df = read_csv("/cloud/project/data/total_para.csv")
tot_df$Prop.para = tot_df$Mummies/tot_df$Total
print(tot_df)
```

Visualising the data:

``` {r visualisation}
ggplot(data = tot_df, aes(x = Treatment, y = Prop.para, 
                          fill = Treatment)) +
  labs(y = "Proportion parasitised") +
  scale_x_discrete(limits=c("NL", "L", "T10", "T50")) +
  geom_boxplot() +
  theme_minimal() +
  scale_fill_viridis(discrete = TRUE, option = "D")
```

It might be better to use a bar graph:

``` {r visualisation 2}
para_prop = group_by(tot_df, Treatment)
by_treatments = summarise(group_by(para_prop, Treatment),
                           mean_proportion = mean(Prop.para),
                           std_dev_proportion = sd(Prop.para)) 
print(by_treatments)

ggplot(data = by_treatments, aes(x = Treatment, y = mean_proportion, fill = Treatment)) +
  labs(y = "Proportion parasitised") +
  scale_x_discrete(limits=c("NL", "L", "T10", "T50")) +
  geom_bar(stat="identity") +
  geom_errorbar(
    aes(ymin = mean_proportion - std_dev_proportion, ymax = mean_proportion + std_dev_proportion),  # Use std_value directly
    width = 0.2, position = position_dodge(0.7))+
  theme_bw() +
  scale_fill_viridis(discrete = TRUE, option = "D")
  
```

Now to analyse the data, using a GLM

```{r analysis}
# Ensure column names are clean
names(tot_df) <- trimws(names(tot_df))

# Calculate proportion parasitism based on surviving aphids
tot_df <- tot_df %>%
  mutate(Prop_Parasitism = Mummies / (Mummies + Surviving))

# Fit a binomial GLM
glm_para <- glm(cbind(Mummies, Surviving) ~ Treatment, 
                family = binomial, data = tot_df)
# Show summary
summary(glm_para)
pairwise_comp = emmeans(glm_para, specs = pairwise ~ Treatment)
summary(pairwise_comp)
```

This was an unexpected result, so I added date as a covariate, to check for any effects it may have 

```{r covariate}

tot_df$Date = as.Date(tot_df$Date, format = "%d.%m") #making sure R recognises this as a date 
tot_df$Date = as.factor(tot_df$Date) #changing into a factor, as each date is a separate experimental day

#now adding date into the model
glm_para <- glm(cbind(Mummies, Surviving) ~ Treatment + Date, 
                family = binomial, data = tot_df) 
summary(glm_para)
pairwise_comp = emmeans(glm_para, specs = pairwise ~ Treatment)
summary(pairwise_comp)

```
There was a higher parasitism rate on 28-01, but a quick check of the data shows that T10 was no underrepresented on that day. Therefore, it's safe to assume that the significant result is not influenced by the date! 

## Total parasitism analysis, HT data
Reading in the data:

``` {r reading HT}
tot_df2 = read_csv("/cloud/project/data/totalHT.csv")
print(tot_df2)

tot_df2$Prop.para = tot_df2$Mummies/tot_df2$Total
```

For this dataset, I was unable to find the survival count for all replicates, so I used the total aphids (in the beginning of the assay) as the denominator for our proportions. 

Visualising the data with boxplots and bar graphs:

``` {r visualisation HT}
#visualising the data:
ggplot(data = tot_df2, aes(x = Treatment, y = Prop.para, 
                          colour = Treatment )) +
  labs(y = "Proportion parasitised") +
  scale_x_discrete(limits=c("CON", "207", "74", "C207", "C74")) +
  geom_boxplot() +
  theme_bw() +
  scale_colour_viridis(discrete = TRUE, option = "D")

#looking at a bar graph
para_prop = group_by(tot_df2, Treatment)
by_treatments2 = summarise(group_by(para_prop, Treatment),
                           mean_proportion = mean(Prop.para),
                           std_dev_proportion = sd(Prop.para))

ggplot(data = by_treatments2, aes(x = Treatment, y = mean_proportion, fill = Treatment)) +
  labs(y = "Proportion parasitised") +
  scale_x_discrete(limits=c("CON", "207", "74", "C207", "C74")) +
  geom_bar(stat="identity") +
  geom_errorbar(
    aes(ymin = mean_proportion - std_dev_proportion, ymax = mean_proportion + std_dev_proportion),  # Use std_value directly
    width = 0.2, position = position_dodge(0.7)) +
  theme_bw() +
   scale_fill_viridis(discrete = TRUE, option = "D")
```

Building a GLM for the HT data:
``` {r GLM HT}
tot_mod2 = lm(Prop.para~Treatment, tot_df2)

glm_para2 <- glm(cbind(Mummies, Total) ~ Treatment, 
                family = binomial, data = tot_df2)

summary(glm_para2)
pairwise_comp2 = emmeans(glm_para2, specs = pairwise ~ Treatment)
summary(pairwise_comp2)

# adding date as a covariate into the model 
tot_df2$Date = as.Date(tot_df2$Date, format = "%d.%m") #making sure R recognises this as a date 
tot_df2$Date = as.factor(tot_df2$Date) #changing into a factor, as each date is a separate experimental day

#now adding date into the model
glm_para2 <- glm(cbind(Mummies, Total) ~ Treatment + Date, 
                family = binomial, data = tot_df2) 
summary(glm_para2)
pairwise_comp2 = emmeans(glm_para2, specs = pairwise ~ Treatment)
summary(pairwise_comp2)
```